### 2021-07-12 response.tex

We note that assuming a Gamma-distributed generation intervals is different from the standard Gamma-chain trick. We note this later in Secion 2:

``Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
Although the Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}, it is different from the generation-interval distribution we get when infectious periods are Gamma distributed, another standard assumption in epidemic modeling (see \cite{roberts2007model} for discussion).''

\revtext{Section 4 doesn't give references to prior work, and section 5 ignores almost all of the literature on the topic of estimating R from incidence data.}

Thank you for the suggestions. 
We have added references in Section 4.
We have also added a new paragraph in the beginning of Section 5 reviewing previous work on estimating R from incidence data:

``Instead of inferring relative strength from speed, one can directly estimate time-varying reproduction numbers $\RR(t)$ of the variant and the wild type from incidence data and directly compare their ratios when the incidence of infection and the generation-interval distributions are known---such methods have been used in previous analyses of the B.1.1.7 strain by \cite{volz2021transmission}.
Broadly, there are two types of reproduction numbers: case reproduction number and instantaneous reproduction number.
The case reproduction number is defined as the average number of secondary infections caused by an individual infected at time $t$ and therefore depends on transmission after time $t$ \citep{wallinga2004different}.
The instantaneous reproduction number is defined as the average number of secondary infections that would be caused by an individual infected at time $t$ if conditions were to remain the same \citep{fraser2007estimating}; 
therefore, the instantaneous reproduction number only depends on transmission at time $t$ and is most appropriate for real-time evaluation of changes in transmission, as argued by \citep{gostic2020practical}.
The estimation of the instantaneous reproduction number was popularized by \cite{cori2013new} and has been widely adopted in epidemiological analyses of SARS-CoV-2 \citep{abbott2020estimating,knight2020estimating,flaxman2020Rt,brauner2021inferring,li2021temporal}.''

@@

\revtext{Other comments:}

\revtext{Topic on which there is a considerable confusion in terminology in the literature.  In addition to "generation interval", terms "generation time", "transmission interval", "serial interval", and others have been used in the literature, all with related by different definitions.  This manuscripts definition "time between infection and transmission" implies a single transmission event, but Kenah pointed out that time-contraction appears when allowing multiple transmission events (see Champredon 2015).  A clearer definition up front will help readers.  Fortunately, the results don't depend on the specific definition.}

Thank you for the suggestions. We have tried to clarify what we mean by the generation interval. In particular, often-menetioned time contractions of generation intervals apply to the forward realized generation interval, which measure time between actual infection events, rather than the intrinsic generation interval, which measure the intrinsic infectiousness of infected individuals. Under strength-like interventions, the intrinsic generation interval does not vary across time, even though the realized generation interval contracts. These distinctions are clearly made in Champredon 2015, and we have tried to clarify these ideas again here in Section 2:

@@

Thank you for the review. We have tried to make the limitation and assumptions clearer throughout text.

\revtext{1. (page 3, first paragraph) It is important to acknowledge that the renewal equation approach to the relationship between the generation interval
distribution, R, and the growth rate r assumes a mass-action epidemic.
When we assume a fixed generation interval distribution, we must add the
assumption that the rate of depletion of susceptibles over time is negligible and the prevalence of infection is low. One of the most important
findings of Ref 10 is that the mean generation interval contracts when the
rate of depletion of susceptibles increases. While the renewal equation
does allow for the generation interval to change over time, the analyses
in Sections 3–5 assume fixed mean generation intervals. Because variants
are likely to spread when susceptibles are being depleted by both variants
(due to cross-protection) and when the prevalence of infection is not low, these implicit assumptions place significant limitations on the practical
implications of the analysis.}

Thank you for pointing this out.
We acknowledge that we are assuming mass action.
However, a fixed generation-interval distribution does not rely on the assumption that the rate of depletion of susceptibles over time is negligible.
This is due to an ongoing confusion between the intrinsic and realized generation interval as we explained above.
For example, as shown in Champredon et al 2015 and 2018, the standard SEIR model can be equivalently written in terms of the renewal equation using a fixed \emph{intrinsic} generation-interval distribution, and we can still observe the contraction of the \emph{realized} generation interval for this model.
We have added a paragraph in Section 2 that makes this distiction clearly (please see response to reviewer 1 for the relevant text).

\revtext{(page 2, lines 55–60) It would help to clarify what is meant by “strengthlike” and “speed-like” in interventions. From the examples, it appears that
speed-like interventions work by identifying potential infections quickly
(effectively ending the infectious period early while not necessarily altering
the hazard of infectious contact) and strength-like interventions work by
reducing the hazard of infectious contact (while not necessarily altering
the duration of infectiousness). A good explanation of these ideas is on
lines 228–231 on page 10.}

### 2021-07-13 supp.tex
lots of edits
done for now, i think

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ry}[1]{\Rx{\mathrm{#1}}}
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
\newcommand{\Rw}{\Rx{\wwwt}}
\newcommand{\Rv}{\Rx{\vvvar}}

@@

In order to model multiple co-circulating strain, we write $i_x(t)$ to represent the incidence of strain $x$ (for $x = 1, \dots, N$).
When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of strain $y$ in the population can be written as:
\begin{equation}
p_y(t) = \frac{i_y(t_0) \exp(r_y t)}{\sum_{x=1}^N i_x(t_0) \exp(r_x t)}.
\end{equation}
Logit transforming $p_y(t)$ then gives:
\begin{align}
\mathrm{logit}(p_y(t)) &= \log\left(\frac{p_y(t)}{1-p_y(t)}\right)\\
&= \log\left(\frac{i_y(t_0) \exp(r_y t)}{\sum_{x \neq y} i_x(t_0) \exp(r_x t)}\right)\\
&= \log(i_y(t_0)) + r_y t - \log \left(\sum_{x \neq y} i_x(t_0) \exp(r_x t)\right)
\end{align}
which is a nonlinear function of time $t$.
Therefore, the growth rate measured at the logit scale is expected to change over time when multiple strains are co-circulating. 
Instead, a multinomial logistic regression model is required to infer relative speed \citep{campbell2021increased}.

@@

We use a two-strain renewal equation to simulate the spread of the variant and the wild type.
Ignoring birth and death, the incidence of infection caused by the variant $\iv$ and the wild type $\iw$ assuming perfect cross immunity is given by:
\begin{align}
\frac{dS}{dt} &= - \iv(t) - \iw(t)\\
\iv(t) &= \Rv(t) \int_{0}^\infty \iv(t-\tau) g_{\mathrm{var}}(\tau) \dtau\\
\iw(t) &= \Rw(t) \int_{0}^\infty \iw(t-\tau) g_{\mathrm{wt}}(\tau) \dtau
\end{align}
where $S$ represents the proportion of susceptible individuals.
We discretize the model at the time scale of 0.025 days as described in \cite{park2021forward} and simulated the model with $\iv(0) = 0.001$ and $\iw(0) = 0.1$.
To allow for smooth changes in $\Rw(t)$ around day 30 and 60, we let:
\begin{equation}
\Rw(t) = 2 \times \left[1 - 0.8 \left(\frac{1}{2} + \frac{\arctan(t-30)}{\pi} \right) \right] \times \left[1 + 1.5 \left(\frac{1}{2} + \frac{\arctan(t-60)}{\pi} \right) \right].
\end{equation}
In this case, $\Rw(t)$ remains around 2 and slowly decreases to $0.4 = 2 \times (1-0.8)$ around day 30, and slowly increases back up to $1 = 0.4 \times (1 + 1.5)$ around day 60.
Finally, we assume $\Rv(t) = \rho \Rw(t)$.
From simulated incidence curves between day 15 and 70, we estimate the instantaneous reproduction number using \eref{rt}.
We ignore incidence before day 15 to remove any potential transient effects.

>>>>

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ry}[1]{\Rx{\mathrm{#1}}}
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
\newcommand{\Rw}{\Rx{\wwwt}}
\newcommand{\Rv}{\Rx{\vvvar}}
\newcommand{\hRw}{\hat{\RR}_{\wwwt}}
\newcommand{\hRv}{\hat{\RR}_{\vvvar}}

@@

In order to model $N$ co-circulating strains, we write $i_x(t)$ to represent the incidence of strain $x$ (for $x = 1, \dots, N$).
When incidence is changing exponentially ($i_x(t) = i_x(t_0) \exp(r_x t)$), the proportion of infections caused by strain $y$ can be written as:
\begin{equation}
p_y(t) = \frac{i_y(t_0) \exp(r_y t)}{\sum_{x=1}^N i_x(t_0) \exp(r_x t)}.
\end{equation}
Logit transforming $p_y(t)$ then gives:
\begin{align}
\mathrm{logit}(p_y(t)) &= \log\left(\frac{p_y(t)}{1-p_y(t)}\right)\\
&= \log\left(\frac{i_y(t_0) \exp(r_y t)}{\sum_{x \neq y} i_x(t_0) \exp(r_x t)}\right)\\
&= \log(i_y(t_0)) + r_y t - \log \left(\sum_{x \neq y} i_x(t_0) \exp(r_x t)\right)
\end{align}
which is a nonlinear function of time $t$.
Therefore, the growth rate measured at the logit scale is expected to change over time when multiple strains are co-circulating. 
Instead, a multinomial logistic regression model is required to infer relative speed (see Supplementary Materials in \citep{campbell2021increased} for derivation).

@@

We use a two-strain renewal equation to simulate the spread of the variant and the wild type in Sections 5--6.
Throughout this section, we assume that the intrinsic infectiousness of the variant is $\rho$ times higher than that of the wild type and that both strains have perfect cross immunity against each other.

Ignoring birth and death, the incidence of infection caused by the variant $\iv$ and the wild type $\iw$ under constant-strength intervention is given by:
\begin{align}
\frac{dS}{dt} &= - \iv(t) - \iw(t)\\
\iv(t) &= \rho  \hRw(t) \frac{S(t)}{N} \int_{0}^\infty \iv(t-\tau) g_{\mathrm{var}}(\tau) \dtau\\
\iw(t) &= \hRw(t) \frac{S(t)}{N}  \int_{0}^\infty \iw(t-\tau) g_{\mathrm{wt}}(\tau) \dtau
\end{align}
where $S$ represents the number of susceptible individuals and $N$ represents the population size.
For this model, the instantaneous reproduction numbers are defined as:
\begin{align}
\Rv(t) &= \rho  \hRw(t)S(t)/N\\
\Rw(t) &=\hRw(t) S(t) /N
\end{align}
In this case, the ratio between the instantaneous reproduction numbers (i.e., the relative strength) remain constant over time:
\begin{equation}
\frac{\Rv(t)}{\Rw(t)} = \rho.
\end{equation}

Likewise, the incidence of infection caused by the variant $\iv$ and the wild type $\iw$ under constant-speed intervention is given by:
\begin{align}
\frac{dS}{dt} &= - \iv(t) - \iw(t)\\
\iv(t) &= \rho  \Rw(0) \frac{S(t)}{N} \int_{0}^\infty \left[\iv(t-\tau) g_{\mathrm{var}}(\tau) \exp \left(- \int_0^\tau h(t-\sigma) \dsigma \right)\right] \dtau\\
\iw(t) &= \Rw(0) \frac{S(t)}{N} \int_{0}^\infty \left[\iw(t-\tau) g_{\mathrm{wt}}(\tau) \exp \left(- \int_0^\tau h(t-\sigma) \dsigma \right)\right] \dtau
\end{align}
where $h$ represents the time-varying hazard of isolation.
Following the definition in Equation~(2) of the main text, the instantaneous reproduction numbers are defined as:
\begin{align}
\Rv(t) &= \rho \Rw(0)  \frac{S(t)}{N} \int_{0}^\infty g_{\mathrm{var}}(\tau) \exp \left(- \int_0^\tau h(t-\sigma) \dsigma \right) \dtau \\
\Rw(t) &= \Rw(0)  \frac{S(t)}{N} \int_{0}^\infty g_{\mathrm{wt}}(\tau) \exp \left(- \int_0^\tau h(t-\sigma) \dsigma \right) \dtau
\end{align}
In this case, the ratio between the instantaneous reproduction numbers (i.e., the relative strength) will not remain constant unless both the wild type and the variant have identical generation-interval distributions.

In Section 5 (Figure 3 in the main text), we rely on the constant-strength model for simplicity.
To allow for smooth decrease and increase in $\Rw(t)$ (and $\Rv(t)$) around day 30 and 60, we let:
\begin{equation}
\Rw(t) = 2 \times \left[1 - 0.8 \left(\frac{1}{2} + \frac{\arctan(t-30)}{\pi} \right) \right] \times \left[1 + 1.5 \left(\frac{1}{2} + \frac{\arctan(t-60)}{\pi} \right) \right].
\end{equation}
In this case, $\Rw(t)$ remains around 2 and slowly decreases to $0.4 = 2 \times (1-0.8)$ around day 30, and slowly increases back up to $1 = 0.4 \times (1 + 1.5)$ around day 60.
We also present the analogous simulation with sharp changes in $\RR$ in Figure S1 using:
\begin{equation}
\Rw(t) = \begin{cases}
2 & t < 30\\
0.4 & 30 \leq t < 60\\
1 & 60 \leq t 
\end{cases}
\end{equation}

In Section 6 (Figure 5 in the main text), we compare the impact of constant-strength and constant-speed models in response to an emerging variant.
For simplicity, we only consider the introduction of interventions and not the lifting of interventions.
In both scenarios, we assume that the infections caused by the wild type is stable at the introduction of interventions (i.e., $\RR=1$ and $r=0$).
In the constant-strength scenario, we have: 
\begin{equation}
\Rw(t) = \begin{cases}
1 & t < 30\\
\theta & 30 \leq t
\end{cases}.
\end{equation}
In the constant-speed scenario, we have: 
\begin{align}
\Rw(0) &= 1\\
h(t) &= \begin{cases}
0 & t < 30\\
\phi & 30 \leq t
\end{cases}.
\end{align}
Both $\theta$ and $\phi$ are chosen such that the post-intervention instantaneous reproduction number of the variant is equal to 0.9 when its mean generation interval is 5 days, as in Figure 4 in the main text.
In both cases, we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\pday$ \citep{davies2021estimated}, $\rw=0\pday$, $\Gw = \Gv = 5\,\textrm{days}$, and $\kappa = 1/5$ \citep{ferretti2020quantifying}.

We discretize the model at the time scale of $\Delta t =0.025\ \textrm{days}$ as described in \cite{park2021forward}.
The initial conditions are set such that infections caused by the variant and the wild type can exhibit exponential changes from $\iv(0) = 0.001$ and $\iw(0) = 0.1$ since day 0.
The population size is set to $N=40,000$
